```{r indi-heur-setup}
library(tidyverse)
library(arrow)
```

# Heuristic Investigation

## Individual Heuristic Investigation

### Rationale

The motivation for the artefact is to investigate advantages ML methods have over existing processes, which make use of heuristics curves and tables detailed in the Halcrow document. In order to use these as a comparison, a research task was designed to faithfully use heuristics to produce high quality predictions. The rationale has several parts:

- Understand specific tasks necessary to make a faithful prediction of condition grade using halcrow information
- Qualitatively understand feasibility of creating a national scale screening tool using halcrow predictions
- Measure performance of heuristic predictions for use as a comparative benchmark when evaluating ML approaches
- Qualitatively understand quantity and depth of expert knowledge required to use halcrow information effectively

It is acknowledged that the author's lack of expertise in flood defence engineering is a major caveat to any predictions made. This is explicitly specified in Halcrow guidance and means results presented here should not be interpreted as the outcome of perfect application. However, undertaking the task with best intentions to learn and apply good judgement will lead to better understanding of what and how much expertise is needed. This speaks directly to the project aims - to provide a way a non-expert responsible for the entire national register of flood defences could benefit from ML tools encoding that expertise.

### Method

The investigation method is detailed in @fig-indi-heuristic-flow. This describes the multi-step process and details actions taken and decisions made to produce predictions. Since it was anticipated this would be intensive and open ended, this was a time-boxed task and initial data was strictly scoped to a manageable subset of the flood defence register at the outset. Where there was any doubt about the outcome of a process, such as material of an asset, the asset was rejected. The focus of this task was to ensure there was enough data to use heuristics without making assumptions about asset characteristics. The final dataset contained only 11 high-confidence assets that were used to make predictions after extensive search of public information sources. All collected data are available in the code repository. 

Once collected, heuristic curves were implemented in a python notebook to predict an asset's current condition grade. Simple metrics were calculated to evaluate success of the method. The implementation is in notebook 13. 

```{=latex}
\begin{adjustwidth}{-2cm}{-2cm}
```
![Individual Asset Heuristic Investigation]("figures/IndividualFloodDefenceAssetHeuristicsPrediction.drawio (1).pdf"){#fig-indi-heuristic-flow width=100%}

```{=latex}
\end{adjustwidth}
```

### Results and Findings

|                                                    | Assets Evaluated | MAE (Discrete ordinal grade) |
|----------------------------------------------------|------------------|----------------------------------------------|
| Individually Assessed Heuristic                    | 11               | 0.272727                                     |
| Individually Assessed Heuristic - Embankments Only | 5                | 0.6                                          |
| National Scale Heuristic                           | 14262            | 0.794041                                     |
| Naive Median Prediction                            | 11               | 0.636364                                     |
| Naive Median Prediction - Embankments Only         | 5                | 0.2                                          |
| Naive Distribution Prediction                      | 11               | 1.272727                                     |
| Naive Distribution Prediction - Embankments Only   | 5                | 0.8                                          |
: Heuristic Prediction Metrics {#tbl-heuristic-pred-metrics}

│ asset_id | pred_grade | actual_grade │
|----------|------------|--------------|
│ 327169   | 2          | 2.0          │
│ 328183   | 2          | 2.0          │
│ 328184   | 2          | 2.0          │
│ 328185   | 2          | 2.0          │
│ 328190   | 2          | 2.0          │
│ 328191   | 2          | 2.0          │
│ 102898   | 3          | 3.0          │
│ 105361   | 3          | 3.0          │
│ 126419   | 3          | 2.0          │
│ 133876   | 3          | 3.0          │
│ 134914   | 5          | 3.0          │
: Heuristic Prediction Results {#tbl-heuristic-predictions}

Predictions using asset-by-asset investigations are promising, with six out of six flood gates (assets with ids beginning 32... , @tbl-heuristic-predictions) assessed predicted perfectly, and overall mean absolute error (MAE) was ~0.27 (@tbl-heuristic-pred-metrics). However, assessed embankments showed MAE of 0.6, with two of assets predicted to be worse than the actual grade. This is interesting for two reasons:

1. Direction of the error means these assets would be flagged as needing attention by a tool, potentially wasting resources
2. Embankments were the asset type requiring most manual work to assess, with multiple sources of imagery and manual elevation profiles drawn.

The main finding is confirmation that producing these assets on a national scale is infeasible, and therefore heuristics cannot be used accurately, with application of local knowledge and domain expertise, to construct a screening tool to help flexible scheduling of inspections because it would take too long to collect data. More importantly, all data collected is fixed, so the exact deterioration rate according to heuristics is known from the point of construction. No extra value can be extracted that would update information and mean that a flexible schedule could improve. This confirms the main advantage of a ML approach - new data can be collected to encode signal from events happening to flood defences during their lifetime or discovered after construction. The reason this investigation is impractical at a national scale is twofold: amount of time taken (several days) to compile trustworthy data for just eleven assets; and high rejection rate, with hundreds of assets screened out before all data necessary to select a curve was collected.

It is acknowledged selection of the correct curve would ordinarily be decided by an expert engineer with local knowledge, but since it was not possible to access this expertise, this investigation's findings support the project's premise that ML approaches hold promise.

## National Scale Implementation

### Rationale

In order to directly compare performance of ML models to heuristics and answer the research question, heuristic predictions must be calculated at comparible scale. Learnings from the Individual Heuristic Investigation confirm infeasibility of producing faithful predictions where each curve input is verified asset-by-asset. The best approach to benchmarking ML models becomes implementation using generalising assumptions to impute data unavailable in the AIMS register or other national datasets. 

By doing so, the number of assets possible to create heuristic estimates for increased to 14,262, approximately 61.2% of assets. 

### Method

Data in the Halcrow document describing construction of heuristics was transcribed from tables into a python script to recreate deterioration curves. The best assumptions for assets having missing information were made, for example all embankments have a variable core material and narrow crest. Implementations of heuristics are found in notebook 8 for original investigation and `flood_defence_condition/src/heuristic.py` for final implementation. Once all inputs are confirmed, the correct curve can be selected and the condition grade is predicted using the asset age. 

Selected heuristic curves are in @fig-selected-heuristic-curves, full curves available in @fig-all-heuristic-curves.


```{r fig-selected-heuristic-curves}
#| fig-cap: "Halcrow deterioration curves for two asset families, by maintenance regime. Each line is one deterioration rate. Applying curves nationally requires selecting rate, which AIMS does not record."
#| fig-width: 5.0
#| fig-height: 5.0

curves <- read_parquet("data/halcrow_curves.parquet") %>%
    filter(
        environment == "fluvial",
        halcrow_type %in% c("embankment", "flood_gate")
    ) %>%
    pivot_longer(
        cols = starts_with("t"),
        names_to = "grade",
        names_prefix = "t",
        values_to = "age"
    ) %>%
    mutate(
        grade = as.integer(grade),
        rate = factor(rate, levels = c("slowest", "medium", "fastest")),
        family = recode_factor(
            halcrow_type,
            embankment = "Embankment, varying core, narrow crest",
            flood_gate = "Flood gate, metal, fluvial environment"
        ),
        regime_label = str_glue("Maintenance regime {regime}")
    )

ggplot(curves, aes(age, grade, colour = rate)) +
    geom_line(linewidth = 0.5) +
    geom_point(size = 0.9) +
    facet_grid(family ~ regime_label, labeller = labeller(family = label_wrap_gen(15))) +
    scale_y_reverse(breaks = 1:5) +
    scale_x_continuous(limits = c(0, 150)) +
    scale_colour_brewer(name = "deterioration rate") +
    labs(x = "asset age (years)", y = "condition grade") +
    theme_minimal(base_size = 9) +
    theme(
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 8)
    )
```

### Results and Findings

The main results of heuristic prediction at scale will be presented in Chapter 7 and be evaluated in comparison with ML results and naive baselines. @tbl-heuristic-pred-metrics contains MAE for immediate comparison with the full asset-by-asset investigation. This shows a large increase in error, demonstrating generalising assumptions don't hold, preventing heuristic application from being effectively scaled. If error of scaled predictions were in line with the first exercise, this would undermine the need for an ML approach. Instead, there are two key findings calling for a new approach to be developed: high error, almost as poor as naive predictors (shown in chapter 7); and lost coverage, with only 62% of assets used in ML experiments able to implement a curve, normally (~97% of missing) because asset age is not trustworth enough to use.